---
title: "Activity 6"
author: "Shannon Duffy"
date: "10/19/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
install.packages(c("raster","sp","rgdal","rgeos","plyr"))
```
```{r}
library(raster)
library(sp)
library(rgdal)
library(rgeos)
library(plyr)
```
```{r}
#read in shapefiles
#readOGR in rgdal does this
g1966 <- readOGR("data6/GNPglaciers/GNPglaciers_1966.shp", stringsAsFactors = T)
g1998 <- readOGR("data6/GNPglaciers/GNPglaciers_1998.shp", stringsAsFactors = T)
g2005 <- readOGR("data6/GNPglaciers/GNPglaciers_2005.shp", stringsAsFactors = T)
g2015 <- readOGR("data6/GNPglaciers/GNPglaciers_2015.shp", stringsAsFactors = T)
```
```{r}
str(g2015)
#data stores all accompanying info/measurements for each spatial object
head(g2015@data)
#polygons stores the coordinates for drawing the polygons
g2015@polygons[[1]]
g1966@proj4string
```
Question 1: Look up the projection and datum. What does zone mean? Describe the projected coordinate system and the properties that it maintains. What spatial scale is this projected coordinate system meant for?

The zone is the 6 degree wedge running north-south along the globe where the cylindrical projection intersects the datum. UTM conserves distances and angles but distorts area toward the poles. This projected coordinate system is meant for small spatial scales where the data you want to examine do not cross multiple zones. 

```{r}
#plot glaciers in 1966
spplot(g1966, "GLACNAME")
```

```{r}
#fix glacier name so that it is consistent with the entire time period
g2015@data$GLACNAME <- ifelse(g2015@data$GLACNAME == "North Swiftcurrent Glacier",
                              "N. Swiftcurrent Glacier",
                                          ifelse(   g2015@data$GLACNAME ==  "Miche Wabun", 
                                               "Miche Wabun Glacier",
                                                as.character(g2015@data$GLACNAME)))
```

```{r}
#read in rgb imagery from landsat
redL <- raster("data6/glacier_09_05_14/l08_red.tif")
greenL <- raster("data6/glacier_09_05_14/l08_green.tif")
blueL <- raster("data6/glacier_09_05_14/l08_blue.tif")
```

```{r}
redL@crs
```
```{r}
#make a brick that stacks all layers
rgbL <- brick(redL, greenL, blueL)
```

```{r}
#plot with color
#show axes for reference
#add contrast to the imagery to see it better
par(mai=c(1,1,1,1))
plotRGB(rgbL, stretch="lin", axes=TRUE)
#add polygons to plot
plot(g1966, col="tan3", border=NA, add=TRUE)
plot(g1998, col="royalblue3", add=TRUE, border=NA)
plot(g2005, col="darkgoldenrod4", add=TRUE, border=NA)
plot(g2015, col="tomato3", add=TRUE, border=NA)
```
```{r}
plotRGB(rgbL, ext=c(289995,310000,5371253,5400000), stretch="lin")
plot(g1966, col="palegreen2", border=NA, add=TRUE)
plot(g1998, col="royalblue3", add=TRUE, border=NA)
plot(g2005, col="darkgoldenrod4", add=TRUE, border=NA)
plot(g2015, col="tomato3", add=TRUE, border=NA)
```
Question 2: Why do you want to work with glacier data in an equal area projection rather than the UTM projection?

Question 3: What happens if you try to plot NDVI with the 1966 glacier polygons? Make a plot of the 2003 NDVI data side by side with the 1966 glacier extent. Explain why you can't put both data files on the same map.

Question 4: Make a map with both the maximum NDVI and the glaciers in 2015. Don't show the axes labels with the x and y coordinate system values. Make the 2015 glacier polygon with no fill color and a black border. What are the patterns in NDVI in the map and around the glaciers?